---
title: "Proteome Quick Analysis"
output:
  html_notebook:
    author: "David McGaughey"
    date: "`r Sys.Date()`"
    theme: flatly
    toc: yes
    toc_float: yes
    code_folding: hide
---


Just checking my ability to load the [AHP data](http://ahp.augusta.edu) and see overall structure with a PCA

```{r load}
library(tidyverse)
library(metamoRph)

proteome_df <- data.table::fread('../data/AhpDb_PsmMatrix.csv')
proteome_df <- proteome_df[-1,] # remove visual formatting line
# make numeric
proteome_matrix <- proteome_df[,-1] %>% as.matrix()
mode(proteome_matrix) = "numeric"
row.names(proteome_matrix) <- proteome_df$Accession
proteome_meta <- read_delim('../data/AhpDb_ProteinSummary.csv', delim = 'asdf') 
colnames(proteome_meta) <- 'one'

# some wacky stuff to handle the ... interesting ... data format
proteome_meta <- proteome_meta %>% 
  mutate_if(is.character, utf8::utf8_encode) %>% 
  separate(one, into = 
             c("UniProt_Id", "Gene_Symbol", "Protein_Name", "Gene_Names", "Accession", "PsmSum", "MeanPsm", "PsmPresent", "PctPresent"), 
           sep = '\\s\\s+,|,\\s\\s+')
# first row is visual formatting
# last row is empty
proteome_meta <- proteome_meta[-c(1,1649),]

sample_meta <- read_csv('../data/AhpDb_ClinicalData.csv')
sample_meta <- sample_meta[-c(1,314),]
```

# PSM Sum Distributions
Should remove the handful of samples with a PSM sum less than 2000 and sum normalize (a la TPM)
```{r}
proteome_matrix %>% as_tibble() %>% select(contains("AH")) %>% colSums(na.rm = TRUE) %>% density() %>% plot()

proteome_matrix %>% as_tibble() %>% select(contains("AH")) %>% colSums(na.rm = TRUE) %>% summary()
```

# Remove low sum PSM samples
```{r}
remove_id <-  proteome_matrix %>% as_tibble() %>% select(contains("AH")) %>% colSums(na.rm = TRUE) %>% enframe() %>% filter(value < 2000)
proteome_tib <- proteome_matrix %>% as_tibble(rownames = 'Accession') %>% select(-contains(remove_id$name))
proteome_tib[is.na(proteome_tib)] <- 0
```

# Check per sample distributions

First 100 samples

VERY long right tail
```{r, fig.height=10, fig.width=5}
proteome_tib[,1:30] %>% pivot_longer(-Accession) %>% 
  ggplot(aes(y=name,x=value)) +
  ggridges::geom_density_ridges2() + 
  cowplot::theme_cowplot()
```
## log1p() Transformed
STILL crazy right tail. Essentially virtually all the data is near zero
```{r, fig.height=10, fig.width=5}
proteome_tib[,1:30] %>% pivot_longer(-Accession) %>% 
  ggplot(aes(y=name,x=log1p(value))) +
  ggridges::geom_density_ridges2() + 
  ylab("Sample") +
  cowplot::theme_cowplot()
```

## log1p() Transformed and remove low counts proteins

Summary of row sums (summing all counts for each protein by sample)
```{r}
proteome_tib[,-1] %>% rowSums(na.rm = TRUE) %>% summary()
```
STILL STILL crazy right tail, even after removing all proteins with a sum < 311 (essentially average 1 count across the 311 samples)
```{r, fig.height=10, fig.width=5}
proteome_tib[proteome_tib[,-1] %>% rowSums(na.rm = TRUE) > 311,1:30] %>% 
  pivot_longer(-Accession) %>% 
  ggplot(aes(y=name,x=log1p(value))) +
  ggridges::geom_density_ridges2() + 
  ylab("Sample") +
  cowplot::theme_cowplot()
```


# PCA
Eight samples are REALLY different than the rest
```{r}
proteome_mat_clean <- proteome_tib %>% select(contains("AH")) %>% as.matrix()
row.names(proteome_mat_clean) <- proteome_tib$Accession
aligned_sample_meta <- colnames(proteome_mat_clean) %>% enframe(value = 'SampleId') %>% select(-name) %>% 
  left_join(sample_meta)
proteome_mat_clean[is.na(proteome_mat_clean)] <- 0
pca_mm <- run_pca(proteome_mat_clean,aligned_sample_meta)

plot <- pca_mm$PCA$x %>% as_tibble(rownames = 'SampleId') %>% 
  left_join(pca_mm$meta, by = 'SampleId') %>% 
  ggplot(aes(x=PC1,y=PC2,color = CD, label = SampleId)) + 
  geom_point() +
  scale_color_manual(values = pals::alphabet() %>% unname()) +
  cowplot::theme_cowplot()
plotly::ggplotly(plot)
```

# PCA Again
Removing those eight samples and AH602 (top left) and redoing the PCA

Looks reasonable now
```{r}
proteome_mat_clean_filter <- proteome_mat_clean[,pca_mm$PCA$x[,c('PC1','PC2')] %>% as_tibble(rownames = 'sample') %>%  filter(PC1 < 20, PC2 < 20) %>% pull(sample)]

aligned_sample_meta <- colnames(proteome_mat_clean_filter) %>% enframe(value = 'SampleId') %>% select(-name) %>% 
  left_join(sample_meta)

pca_mm <- run_pca(proteome_mat_clean_filter,aligned_sample_meta)

plot <- pca_mm$PCA$x %>% as_tibble(rownames = 'SampleId') %>% 
  left_join(pca_mm$meta, by = 'SampleId') %>% 
  ggplot(aes(x=PC1,y=PC2,color = CD, label = SampleId)) + 
  geom_point() +
  scale_color_manual(values = pals::alphabet() %>% unname()) +
  cowplot::theme_cowplot()
plotly::ggplotly(plot)
```


# Output cleaned data
```{r}
save(proteome_mat_clean_filter, aligned_sample_meta, proteome_meta, file = '../data/ahpdb_cleaned_data.Rdata')
```

# Session Info
```{r}
sessionInfo()
```
